Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-08-17 ms.Rnw
Ed's comments

However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We write $r^u=(r^u_i)_{i\in\Demes}$.
Note that we lineages that die as a result of an event do not count in the production.
Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $t\in\Rp$,
  $Y^{P}(t)=\left(\ct^{P}(t),\anc^{P}(t),\deme^{P}(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
Given $y\in\Yspace$, we will use the notation $\ct(y)$, $\anc(y)$, and $\deme(y)$ to refer to the coordinates of $y$.

@@

\begin{thm}\label{thm:pruned_lik}
  Let $P^*_t$ be a given pruned genealogy and $y^*=Y^{P^*_t}$.
  Then
  \begin{equation*}
    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}\,Q_{U_e}(\yt^*_e,y^*_e)}
    \:\times\:\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
  \end{equation*}
\end{thm}
\begin{proof}
  As we have already observed, if $\event{P^*}{t}\nsubseteq\event{\Hist}{t}$, then $\Prob{P^*_t\vert\Hist_t}=0$.
  Similarly, if there is any event of $P^*_t$ which is incompatible with the population process, $\Prob{P^*_t\vert\Hist_t}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t)$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*_t$.
  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*_t\vert\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}}.
  \end{equation*}
  Returning to the possibility that $P^*_t$ is incompatible with $\Hist_t$, since $\Prob{P^*_t}=0$ if either any $Q_u=0$ or any event of $P^*_t$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Let $V^*_T$, $T>0$ be a given obscured genealogy.
  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
    \end{gathered}
  \end{equation*}
  Suppose $w=w(t,x)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &\ +\sum_{e\in\event{V^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)$.
  Then the likelihood of $V^*_T$ is
  \begin{equation*}
    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
\end{thm}

@@

\begin{defn}
  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
  \end{equation}
  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x)\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
    \end{aligned}
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
\end{defn}

@@

\begin{lemma}\label{lemma:filtereq}
  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
    \end{aligned}
  \end{equation}
  on the space $\Xspace\times(0,\infty)$.
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
\end{lemma}
\begin{proof}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

@@

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a non-explosive pure jump process with KFE
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      w(0,x)=p_0(x).
    \end{gathered}
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{V}_t$ is a real-valued random process such that
  \begin{equation}\label{eq:Vprod}
    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{B(e,\leftlim{X}_e,X_e)},
  \end{equation}
  for some given measurable function $B\ge{0}$.
  If $w$ satisfies the filter equation,
  \begin{equation*}
    \frac{\partial{w}}{\partial{t}}(t,x)
    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
  \end{equation*}
  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
  \begin{equation*}
    \begin{gathered}
      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
      \qquad
      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
    \end{gathered}
  \end{equation*}
\end{proof}

@@

\begin{prop}\label{prop:fullfilt}
  Let $\X_t$, $\Hist_t$, $B$, and $\rdm{V}_t$ be as in \cref{prop:zeroboost}.
  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
  Let $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}$ denote the density of $\rdm{V}_t$ conditional on $\mathcal{M}\subseteq\event{\Hist}{t}$.
  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
    \end{aligned}
  \end{equation*}
  then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
\end{prop}
\begin{proof}
  Suppose $s\in[0,t]$ and let
  \begin{equation*}
    w(s,x)=\Expect{\prod_{e\in\event{\Hist}{s}}{B\left(e,\Xt_e,\X_e\right)}}.
  \end{equation*}
  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
  \begin{equation*}
    \begin{aligned}
      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  Taking the limit as $\Delta\downarrow{0}$, we obtain
  \begin{equation*}
    \frac{\partial{w}}{\partial{s}}(s,x)
    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
  \end{equation*}
  If $s\in\mathcal{M}$, then
  \begin{equation*}
    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
  \end{equation*}
  By \cref{rem:equivfilt}, the result follows.
\end{proof}

>>>>

However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We write $r^u=(r^u_i)_{i\in\Demes}$.
Note that the lineages that die as a result of an event do not count in the production.
Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.

@@

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  \ELI{[How is this defined when there is immigration in the system?]}
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $t\in\Rp$,
  $Y^{P}(t)=\left(\ct^{P}(t),\anc^{P}(t),\deme^{P}(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $y:\Rp\to\Yspace$ defines a pruned genealogy.
If $y$ does define a pruned genealogy, we denote that pruned genealogy by $P(y)$.
Given $y\in\Yspace$, we will use the notation $\ct(y)$, $\anc(y)$, and $\deme(y)$ to refer to the coordinates of $y$.

@@

\begin{thm}\label{thm:pruned_lik}
  Let $P^*_t$ be a given pruned genealogy and $y^*=Y^{P^*_t}$.
  Then
  \begin{equation*}
    \Prob{P^*_t\vert\Hist_t}=\prod_{e\in\event{\Hist}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}\,Q_{U_e}(\yt^*_e,y^*_e)}
    \:\times\:\Indicator{\event{P^*}{t}\subseteq{\event{\Hist}{t}}}.
  \end{equation*}
  \ELI{[Do we need $Q$, if the binomial ratio is defined to be zero in impossible cases?]}
\end{thm}
\begin{proof}
  As we have already observed, if $\event{P^*}{t}\nsubseteq\event{\Hist}{t}$, then $\Prob{P^*_t\vert\Hist_t}=0$.
  Similarly, if there is any event of $P^*_t$ which is incompatible with the population process, $\Prob{P^*_t\vert\Hist_t}=0$.
  Let us therefore suppose that neither of these conditions hold.
  Recall that, conditional on $\Hist_t$, each jump of the inventory process is independent of the others.
  Moreover, at each event $e\in\event{\Hist}{t}$, a jump of mark $U_e$ occured, with a production of $r^{U_e}=(r_i)_{i\in\Demes}$, resulting in a new deme-occupancy of $n(X_e)=(n_i)_{i\in\Demes}$.
  In $P^*_t$, at time $e$, there are $\ell_i=\ell_i(y^*_t)$ lineages in deme $i$, of which $s_i=s_i(\yt^*_t,y^*_t)$ are emergent.
  The exchangeability of lineages within demes implies that each lineage present in a deme at time $e$ was equally likely to have been one of the emergent lineages.
  In particular, at time $e$, the probability that $s_i$ of the $\ell_i$ deme-$i$ lineages were among the $r_i$ of $n_i$ lineages emergent in the inventory process is the same as the probability that, upon drawing $\ell_i$ balls without replacement from an urn containing $r_i$ black balls and $n_i-r_i$ white balls, exactly $s_i$ of the drawn balls are black, namely
  \begin{equation*}
    \frac{\binom{n_i-\ell_i}{r_i-s_i}\,\binom{\ell_i}{s_i}}{\binom{n_i}{r_i}}.
  \end{equation*}
  Because our lineages are labelled, each of the $\binom{\ell_i}{s_i}$ equally probable sets of $s_i$ lineages is distinct; just one of these is the one present in $P^*_t$.
  Moreover, since, again conditional on $\Hist_t$, the inventory process within each deme is independent of the others, we have established that
  \begin{equation*}
    \Prob{P^*_t\vert\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{\BinRatio{n(X_e)}{\ell(y^*_e)}{r^{U_e}}{s(\yt^*_e,y^*_e)}}.
  \end{equation*}
  Returning to the possibility that $P^*_t$ is incompatible with $\Hist_t$, since $\Prob{P^*_t}=0$ if either any $Q_u=0$ or any event of $P^*_t$ is not an event of $\Hist_t$, we obtain the result.
\end{proof}

@@

\begin{thm}
  Let $V^*_T$, $T>0$ be a given obscured genealogy.
  Let $\pi$, $\kappa_{V^*_T}$, and $\rdm{y}_t$ be as in \cref{thm:obsc_lik}.
  For $x,x'\in\Xspace$, $y,y'\in\Yspace$, $u\in\Jumps$, define
  \begin{equation*}
    \begin{gathered}
      \phi_u\big(t,x,y,y')=\BinRatio{n(x)}{\ell(y')}{r^u}{s(y,y')}\,\frac{\kappa_{V^*_T}(t,y,y')}{\pi_u(t,x,x',y,y')},\\
      \beta_u(t,x,x',y,y')=\alpha_u(t,x,x')\,\pi_u(t,x,x',y,y').
    \end{gathered}
  \end{equation*}
  Suppose $w=w(t,x,y)$ satisfies the filter equation
  \begin{equation}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x,y)=
      &\sum_{uy'}{\int{w(t,x',y')\,\beta_u(t,x',x,y',y)\,\phi_u(t,x,y',y)\,\dd{x'}}}\\
      &\ -\sum_{uy'}{\int{w(t,x,y)\,\beta_u(t,x,x',y,y')\,\dd{x'}}}\\
      &\ +\sum_{e\in\event{V^*}{T}}{\delta(t,e)\,\left\{\sum_{uy'}{\int{w(t,x')\,\beta_u(t,x',x)\,\phi_u(t,x,y',y)\,\dd{x'}}}-w(t,x)\right\}},
    \end{aligned}
  \end{equation}
  with the initial condition $w(0,x,y)=p_0(x)$.
  Then the likelihood of $V^*_T$ is
  \begin{equation*}
    \lik(V^*_T)=\sum_y{\int{w(T,x,y)\,\dd{x}}}.
  \end{equation*}
  \ELI{[I was surprised by the assertion of existence of $k_V$ here, and then the proof providing a specific construction---why not assert the exact form?
      The proof is also unusual---via looking for an importance sampling solution.]}
\end{thm}

@@

\begin{defn}
  Suppose $\X_t$ is a continuous-time Markov process with Kolmogorov forward equation (KFE)
  \begin{equation}\label{eq:kfe2}
    \frac{\partial{w}}{\partial{t}}
    =\int{w(t,x')\,\beta(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}.
  \end{equation}
  Suppose that $B(t,x,x')\ge{0}$ and $\lambda(t,x)\in\R$ are given measurable functions.
  \ELI{[Ranges are defined elementwise, which is not the usual math formalism,
      e.g., $\lambda:T\times\Xspace\to\mathbb{R}$
      or just say $\lambda(t,x)$ takes values in $\mathbb{R}$.]}
  Let $\mathcal{M}\subseteq\Rp$ be countable such that $\mathcal{M}\cap[0,t]$ is finite for all $t$.
  We say that the equation
  \begin{equation}\label{eq:filtereq}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\beta(t,x,x')\,\dd{x'}}
      -\lambda(t,x)\,w(t,x)\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\beta(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\}.
    \end{aligned}
  \end{equation}
  is the \emph{filter equation} with \emph{driver} $\X_t$ (or \emph{generated by} $\beta$), \emph{boost} $B$, \emph{decay} $\lambda$, and \emph{observed events} $\mathcal{M}$.
  \ELI{[Why is $B=1$, $\lambda=0$ not equal to Eq.~(A1)?
      (It is if we have $|M|=0$.)]}
  \AAK{[The KFE is a special case of a filter equation.]}
\end{defn}

@@

\begin{lemma}\label{lemma:filtereq}
  The filter equation \eqref{eq:filtereq} with $\mathcal{M}=\emptyset$ and $B>0$ is satisfied by $w(t,x)=\int_0^{\infty}{v\,u(t,x,v)\,\dd{v}}$, where $u(t,x,v)$ satisfies the KFE
  \begin{equation}\label{eq:filterlemma}
    \begin{aligned}
      \frac{\partial{u}}{\partial{t}}=&\int{u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)\,v')\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')\,v)\,\dd{x'}\,\dd{v'}}
      +\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right],
    \end{aligned}
  \end{equation}
  on the space $\Xspace\times(0,\infty)$.
  Here, $\delta(v,v')$ is the familiar Dirac $\delta$.
  \ELI{[Note: the last term in eq (A3) is not in the form of $-lambda\,u$ for some $\lambda$.
      So, Eq (A3) is not a filter equation in the form of Eq (A1).
  ]}
  \AAK{[But it is a KFE, for a process with a drift.]}
\end{lemma}
\begin{proof}
  \ELI{[Conditions for differentiating under the integral in the proof?
      Need $[v^2 \lambda u(t,x,v)]_{v=0}^{\infty} = 0$.
      That's okay at $v=0$, and a minor constraint on $u$.]}
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v\,\frac{\partial{u}}{\partial{t}}(t,x,v)\,\dd{v}}\\
      =&\int{v\,u(t,x',v')\,\beta(t,x',x)\,\delta(v,B(t,x',x)v')\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad-\int{v\,u(t,x,v)\,\beta(t,x,x')\,\delta(v',B(t,x,x')v)\,\dd{v}\,\dd{x'}\,\dd{v'}}\\
      &\qquad+\int{v\,\tfrac{\partial}{\partial{v}}\left[\lambda(t,x)\,v\,u(t,x,v)\right]\,\dd{v}}.\\
    \end{aligned}
  \end{equation*}
  Evaluating the first integral with respect to $v$, the second with respect to $v'$, and the third by parts, we obtain
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{v'\,u(t,x',v')\,\beta(t,x',x)\,B(t,x',x)\,\dd{v'}\,\dd{x'}}
      -\int{v\,u(t,x,v)\,\beta(t,x,x')\,\dd{v}\,\dd{x'}}\\
      &\qquad-\lambda(t,x)\,\int{v\,u(t,x,v)\,\dd{v}},
    \end{aligned}
  \end{equation*}
  which is simplified to obtain \cref{eq:filtereq}.
\end{proof}

@@

\begin{prop}\label{prop:zeroboost}
  Suppose $\X_t$ is a non-explosive pure jump process with KFE
  \begin{equation*}
    \begin{gathered}
      \frac{\partial{w}}{\partial{t}}(t,x)=\int{w(t,x')\,\alpha(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      w(0,x)=p_0(x).
    \end{gathered}
  \end{equation*}
  Let $\Hist_t$ be its history process.
  Suppose $\rdm{V}_t$ is a real-valued random process such that
  \begin{equation}\label{eq:Vprod}
    \Expect{\rdm{V}_t\;\vert\;\Hist_t=H_t}=\prod_{e\in\event{H}{t}}{B(e,\leftlim{X}_e,X_e)},
  \end{equation}
  for some given measurable function $B\ge{0}$.
  If $w$ satisfies the filter equation,
  \begin{equation*}
    \frac{\partial{w}}{\partial{t}}(t,x)
    =\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
    -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
  \end{equation*}
  then $\Expect{\rdm{V}_t}=\int{w(t,x)\,\dd{x}}$.
  \ELI{[$dw/dt$ for $X$ could be $du/dt$ since this is not the same $w$ as the subsequent line.
      The terminology uses the letter $w$ as a dummy variable, but this makes it harder to talk about the solutions to different KFEs.]}
\end{prop}
\begin{proof}
  Define $Q(t,x,x')=\Indicator{B(t,x,x')>0}$ and apply \cref{lemma:filtereq} with
  \begin{equation*}
    \begin{gathered}
      \beta(t,x,x')=\alpha(t,x,x')\,Q(t,x,x'),
      \qquad
      \lambda=\int{\alpha(t,x,x')\,\left(1-Q(t,x,x')\right)\dd{x'}}.
    \end{gathered}
  \end{equation*}
  \ELI{[Lemma A1 has $B>0$, which only covers the cases when $Q=1$ and $\lambda=0$.
      But, $B\ge{0}$ seems enough in the proof of lemma A1, as long as one allows a one-sided Dirac $\delta$ at $0$.
      Maybe the Dirac $\delta$ should all be one-sided in order to integrate
      correctly on right-continuous functions? ]}
\end{proof}

@@

\begin{prop}\label{prop:fullfilt}
  Let $\X_t$, $\Hist_t$, $B$, and $\rdm{V}_t$ be as in \cref{prop:zeroboost}.
  Suppose that $\mathcal{M}\subseteq\halfclosed{0,t}$ is a fixed, finite set of observed event times.
  Let $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}$ denote the density of $\rdm{V}_t$ conditional on $\mathcal{M}\subseteq\event{\Hist}{t}$.
  If $w(t,x)$ satisfies $w(0,x)=p_0(x)$ and the filter equation
  \begin{equation*}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}
      =&\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}
      -\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}\\
      &\quad+\sum_{e\in\mathcal{M}}\delta(t,e)\,\left\{\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-w(t,x)\right\},
    \end{aligned}
  \end{equation*}
  then $\Expect{\rdm{V}_t\;\vert\;\mathcal{M}}=\int{w(t,x)\,\dd{x}}$.
  \ELI{[(a) line 2: density -> expectation.
      (b) the term $\delta(t,e)\,w(t,x)$ seems irrelevant:
      the derivative at a finite set of points only matters when it is a delta function.
      (c) what is a ``fixed, finite set of event times''?
      $M_t^H$ is a random variable, so it cannot usually be fixed.
      $M$ is also treated sometimes as a random variable, even though it is defined to be fixed.
      I see this means, ``conditional on the event that $M_t^H$ contains $M$'', but we should be explicit, to avoid confusion about whether $M$ is a random variable or not.]}
\end{prop}
\begin{proof}
  Suppose $s\in[0,t]$ and let
  \begin{equation*}
    w(s,x)=\Expect{\prod_{e\in\event{\Hist}{s}}{B\left(e,\Xt_e,\X_e\right)}}.
  \end{equation*}
  That is, $w(s,x)$ is the expectation (over histories) of the partial products of \cref{eq:Vprod}.
  If $s\notin{\mathcal{M}}$, then for $\Delta>0$ sufficiently small,
  \begin{equation*}
    \begin{aligned}
      w(s,x)=&\left(1-\Delta\,\int{\alpha(s-\Delta,x,x')\,\dd{x'}}\right)\,w(s-\Delta,x)\\
      &\qquad+\int{\alpha(s-\Delta,x',x)\,B(s-\Delta,x',x)\,w(s-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  Taking the limit as $\Delta\downarrow{0}$, we obtain
  \begin{equation*}
    \frac{\partial{w}}{\partial{s}}(s,x)
    =\int{w(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}
    -\int{w(s,x)\,\alpha(s,x,x')\,\dd{x'}}.
  \end{equation*}
  If $s\in\mathcal{M}$, then
  \begin{equation*}
    w(s,x)=\int{\wt(s,x')\,\alpha(s,x',x)\,B(s,x',x)\,\dd{x'}}.
  \end{equation*}
  By \cref{rem:equivfilt}, the result follows.
  \ELI{[The argument shows that the filter equation satisfies a KFE, but the limit expression seems justified by a similar KFE.
      Is the point that one KFE implies the other?]}
\end{proof}

\ELI{[Maybe, Prop A3 could be a version of Prop A2 allowing for discrete transitions at a finite set of times;
    a combination of cts time and discrete time Markov chains.
    That is all we need to extend Prop A2 to observation times, which can be viewed just as an extended Markov process including the observation events.
    That is, to make a distinction between ``observation events'' that give the sequences used to construct the phylogeny, and ``observed events'' which are determined points on the phylogeny.
    The latter are a realization of a random process, and the former are defined to be fixed a priori.]}
